---
title: "ADVANZ-4 status report"
author: "Carlos Blázquez Bondia"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(metar)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
library(lme4)
library(lmerTest)
library(vegan)
library(gridExtra)
library(ggrepel)
source(file = here::here("code","myFunctions.R"))
source("code/Paired_wilcox_NA_Handler.R")


```


# Sequencing throupghput analysis and Gene Richness

The median sequencing sample throughput was 20789049 reads, with a quasi-normal distribution. Samples were rarefied to the 98th percentile, which accounted to a threshold of 11888709 (Figure1). 6 samples were below this threshold and hence were removed.


```{r load_metadata_vars, eval=T, include=F}
# mymre<-readRDS(here::here(myPath,"data","mre.rds"))




mymre <- aws.s3::s3readRDS("s3://mistral-wp6-advanz4/metagenome/WMGS/MREObject.rds") ### Check consistency between mre.rds(local and S3). You are reading s3 file, here, but in 200_ you are using local
# metadata <- metar::get_metadata(mymre)@metadata_df

metadata <- here::here("Metadata", "2022_09_14_clean_metadata_LIMS.csv") %>% 
  read.csv() %>% 
  column_to_rownames("SampleID") %>% 
  mutate(SampleID = rownames(.))

mymre@metadata@categorical_vals <-  
  here::here("Metadata","CategoricalVariables.txt") %>%
  read.table(., sep="\t") %>%
  setNames(c("CategoricalVariable","PaletteName")) %>%
  as_tibble()

mymre@metadata@metadata_df <-  tibble(metadata)


cat_df <- 
  here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T)

cat_vector <- cat_df %>% 
  pull(CategoricalVariable)

long_var <- here::here("Metadata", "LongitudinalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(LongitudinalVariable)

link_var <- here::here("Metadata", "LongitudinalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(LinkVariable)

num_var <- here::here("Metadata", "NumericalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(NumericalVariable)


```

```{r update_mre, warning= F, message= F, echo =F}

#### This chunk is for updating the mre internally as new variables come out. Once the final mre is done this chunk will be removed.
devtools::load_all("../WMGSPipeline/")


mymre <- filter_samples(mre = mymre, sample_ids = metadata$SampleID)

mymre@metadata@metadata_df <- as_tibble(metadata)

mymre@metadata@categorical_vals <- here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@numeric_vals <- here::here("Metadata", "NumericalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@longitudinal_vals <- here::here("Metadata", "LongitudinalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()


```



```{r get_GeneRichness_Data, echo=FALSE, message=F, warning=F}


igc<-metar::get_diversity(mymre)@igc

# We'll save this number to report the minimum read threshold.

myigc <- igc_lmm(mymre)



Threshold <-  igc@dataTable %>%
  dplyr::select(SampleID, NumberMappedReads) %>%
  unique() %>%
  summarise(Q = quantile(NumberMappedReads,0.02)) %>% 
  as.numeric()


readCountDistribution<-igc@dataTable %>%
    select(SampleID, NumberMappedReads) %>% 
    unique() %>%

    ggplot(aes(x=NumberMappedReads)) +
    geom_density(fill="grey") + 
    theme_bw() +
    geom_vline(xintercept = Threshold)


TableList <- cat_vector %>%
    set_names() %>%
    map( ~{

        
          ThresholdTable <- igc@dataTable %>%
                filter(NumberMappedReads >= Threshold) %>%
                filter(ReadCountReal>=Threshold) %>% ### Get the Mapped Read count (per-sample) just above the threshold
                group_by(SampleID) %>%
                summarise(across(everything(), min)) %>%
                ungroup() %>%
          right_join(., metadata, by="SampleID") %>%
          select(SampleID,
                 richness = GeneNumber,
                 link_var=link_var,
                 cat_var = .x,
                 long_var = long_var,
                 NumberMappedReads)
  

        
    })

```

```{r get_phyloseqs}
phy <- get_phyloseq(mymre, type="metaphlan")

```


```{r Get_DistMats, eval=T, include=F, cache=T}

getAbundances<-function(phyloseq, level){
  
  collapsedPhyloseq<- phyloseq::tax_glom(phyloseq, level)
  
  OtuTable<- phyloseq::otu_table(collapsedPhyloseq) %>%
    magrittr::set_rownames(as.character(phyloseq::tax_table(collapsedPhyloseq)[,level])) %>%
    as.data.frame()
  return(OtuTable)
  
}




myAbundanceDF<- getAbundances(phy, "Species")



myNMDS <- myAbundanceDF %>%
  t() %>% 
    as.data.frame() %>%
    # vegan::vegdist(., method = "bray") %>%
 
    metaMDS(.,trymax=800,  distance = "bray",
            k = 2) %>%
    pluck("points") %>%
    as.data.frame() 

  # NMDS = data.frame(NMDS1 = dim_NMDS$points[,1], NMDS2 = dim_NMDS$points[,2])

BiPlotCoords<- envfit(myNMDS, t(myAbundanceDF), perm=999) 

BiPlotCoordsDF <- c("MDS1","MDS2","r","pvals") %>%
  set_names() %>%
  
  map_dfr(., ~ {
    
    if(.x %in% c("MDS1","MDS2")){
      BiPlotCoords$vectors$arrows[,.x] %>%
        as.numeric()
    } else{
      BiPlotCoords$vectors %>%
        pluck(.x) %>%
        as.numeric()
    } 
      
  }) %>%
  as.data.frame() %>%
  set_names(nm=c("NMDS1","NMDS2","r","p")) %>%
  dplyr::mutate(Dim1 = NMDS1*sqrt(r),
                Dim2 = NMDS2*sqrt(r),
                species = rownames(myAbundanceDF)) 
  

myBrayDistMat <-vegdist(wisconsin(sqrt(t(myAbundanceDF))))


```




```{r NMDS_Group, warning= F, message= F}

cat_df <- cat_df %>%
  filter(!CategoricalVariable == "CD4diff_96")
cat_var <- "group"



link_var <- "record_id"
id_var <- "SampleID"

myPal<-RColorBrewer::brewer.pal(name =  cat_df$PaletteName[1],
                                n=length(unique(metadata[,cat_var])))
BiPlotCoordsDF<-BiPlotCoordsDF %>%
  mutate(Dim1s = Dim1*3,
         Dim2s = Dim2*3)


myNMDS_DF <- myNMDS %>%
  rownames_to_column(var = id_var) %>%
  dplyr::full_join(., metadata, by =id_var) %>%
  dplyr::select(SampleID = id_var,
                NMDS1 = MDS1,
                NMDS2 = MDS2,
                linkVar = !!sym(link_var),
                catVar = !!sym(cat_var),
                LongVar = long_var)


  myXmin<-min(myNMDS$MDS1)
  # +(max(dim_NMDS$points[,1])-min(dim_NMDS$points[,1]))/20
  myYmin<-max(myNMDS$MDS2)-(max(myNMDS$MDS1)-min(myNMDS$MDS1))




myAdonis<- adonis(myBrayDistMat~ catVar, data = myNMDS_DF) %>%
  pluck("aov.tab") %>%
  as.data.frame() %>%
  slice(1L) %>%
  magrittr::set_rownames(cat_var)


myNMDSplot<-ggplot(myNMDS_DF, aes(x=NMDS1, y=NMDS2)) +
    geom_point() +
    theme_bw()+
    theme(panel.border=element_blank(),
          panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),

          axis.line=element_line(colour="black"),
          axis.text.x=element_text(size=11),
          axis.text.y=element_text(size=11),
          axis.title.x=element_text(size=11),
          axis.title.y=element_text(size=11),

          legend.text=element_text(size=11,face="italic"),
          legend.title=element_text(size=11))+

    ggtitle(expression(atop("Species composition",atop(italic("all samples"))))) +
    theme(plot.title=element_text(lineheight=1,face="bold",size=13))+
    coord_fixed()+
    labs(fill = cat_var, color=cat_var, shape = long_var)




myNMDSBiplot<-ggplot(myNMDS_DF, aes(x=NMDS1, y=NMDS2)) +
    geom_point(aes(colour=catVar, shape = as.character(LongVar))) +
    scale_colour_manual(values = myPal, labels = c("DGT", "DRV/r")) +
    scale_fill_manual(values=myPal, labels = c("DGT", "DRV/r"))+
    theme_bw()+
    theme(panel.border=element_blank(),
          panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),

          axis.line=element_line(colour="black"),
          axis.text.x=element_text(size=11),
          axis.text.y=element_text(size=11),
          axis.title.x=element_text(size=11),
          axis.title.y=element_text(size=11),

          legend.text=element_text(size=11,face="italic"),
          legend.title=element_text(size=11))+

    ggtitle(expression(atop("Species composition",atop(italic("all samples"))))) +
    theme(plot.title=element_text(lineheight=1,face="bold",size=13))+
    stat_ellipse(geom="polygon",
                 alpha=0.15,color="black",
                 aes(fill=catVar),level=0.95)+
    # scale_size_manual(values = c(1:length(unique(myNMDS_DF[,"LongVar"])))/1.5) +
    geom_segment(data=BiPlotCoordsDF,aes(x=0,xend=Dim1s,y=0,yend=Dim2s),
                 arrow = arrow(length = unit(0.2,"cm")),colour="darkgrey")+
    geom_text_repel(data=BiPlotCoordsDF,aes(Dim1s,Dim2s,label=species),
                    size=2.5, box.padding = unit(0.25, "lines"),
                    segment.colour="black",
                    segment.alpha=0.5,
                    segment.size=0.10)+
    coord_fixed()+
    labs(fill = cat_var, color=cat_var, shape = long_var) +
  annotation_custom(gridExtra::tableGrob(round(myAdonis,3),
                                         theme = ttheme_default(base_size = 8)),ymin = 3.5, xmin=2.4)


```



```{r wilcox_across_group, warning=F, message=F, cache = T}

phy <- get_phyloseq(mymre, type = "metaphlan") %>% 
      phyloseq::filter_taxa(., function(x) sum(x > 0) > (0.3*length(x)), TRUE)


otu_df <- 
  phyloseq::otu_table(phy) %>% 
  t() %>% 
  as.data.frame() %>% 
  stats::setNames(phyloseq::tax_table(phy)[,"Species"]) %>% 
  tibble::rownames_to_column("SampleID") %>% 
  stats::setNames(.,stringr::str_remove_all(colnames(.), "\\[")) %>% 
  stats::setNames(.,stringr::str_remove_all(colnames(.), "\\]"))


stat_list <-
  cat_vector %>%
  purrr::set_names() %>%
  purrr::map(function(cv) {
    
    c(0,24,48,96) %>% 
      set_names() %>% 
      purrr::map_dfr(function(tp){
        otu_df %>%
        dplyr::select(-SampleID) %>%
        colnames() %>%
        purrr::set_names() %>%
        purrr::map_dfr(function(sp) {
          df <-
            otu_df %>%
            dplyr::select(SampleID, !!sym(sp)) %>%
            dplyr::left_join(metadata[, c("SampleID", link_var, cv, long_var), drop = F], 
                             by = "SampleID") %>% 
            dplyr::filter(!!sym(long_var) %in% tp) 

          
          if (length(unique(dplyr::pull(.data = df, !!sym(cv)))) == 2) {
            test <- df %>%
              rstatix::wilcox_test(., formula(paste(sp, "~", cv)), paired = F) %>%
              dplyr::mutate(method = "Wilcoxon") 
            
          } else{
            test <- df %>%
              rstatix::kruskal_test(., formula(paste(sp, "~", cv))) %>%
              dplyr::mutate(method = "Kruskal-Wallis")
          }
          
        }) %>% 
      rename(species = .y.) %>% 
      dplyr::mutate(long_var = tp) %>% 
      group_by(long_var) %>% 
      mutate(p.adj = p.adjust(p, "BH")) %>% 
      # filter(p.adj < .05 ) %>% 
      group_by(species) 
      }) 
}) 


DAsp <- stat_list$group %>% 
  filter(p<0.05) %>% 
  group_by(species)


## --- init MNJ
## It is interesting that Bifidobacterium species (adolescentis, longum) are consistently found as DA in all (most) timepoints
## Produce a table showing 
sp_detected <- 
stat_list$group %>% 
  filter(p<0.05) %>% 
  group_by(species) %>% 
  tally() %>% 
  filter(n>2) %>%
  arrange(desc(n))## Species being significant (Wilcoxon unadjusted <0.05)

sp_detected




  
### Show how many species are differential in at least two of the timepoints in each cat var. Risk_group should result in many species showing that kins of signal. See Venn section below.
### --- end MNJ

```


Bifidobacterium adolescentis, Bifidobacterium longum and Olsenella scatoligenes were significant in all timepoints. Let's check the actual abundances per group and timepoint:


```{r plot_DA, echo = F, message = F, warning=F, eval=F, include=F}
 c("Olsenella_scatoligenes", "Bifidobacterium_adolescentis", "Bifidobacterium_longum") %>% 
  purrr::map(function(sp){
  otu_df %>% 
  select(SampleID, species = !!sym(sp)) %>% 
  # select(SampleID, !!sym(DAsp)) %>% 
  left_join(., metadata[,c("SampleID","group","time_point")], by="SampleID") %>% 
  mutate(time_point = as.factor(time_point)) %>%     
  # ggstatsplot::ggbetweenstats(., x="group",y="species")
  # group_by(group) %>%
  ggstatsplot::grouped_ggbetweenstats(data = .,
                              x=time_point,
                              y = species,
                              # type="np",
                              grouping.var = group,
                              # bf.message=NULL,
                              # ylab = paste(sp,"(%)", sep = ""), 
                              # ggtheme = ggplot2::theme_bw(),
                              plotgrid.args = NULL)


  }) %>% 
  ggpubr::ggarrange(plotlist = ., nrow = 3)
```

Error in unique.default(x, nmax = nmax) : 
unique() applies only to vectors
