---
title: "BL_stratification"
author: "Carlos Blázquez Bondia"
date: "2023-03-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
source(here::here("code","LMM.R"))
source(here::here("code","myFunctions.R"))
library(RColorBrewer)

```

```{r load_data, message=F, warning=F, echo=FALSE}

mymre <- here::here("data", "filt_mre.rds") %>% 
  readRDS()


metadata <-
  here::here("Metadata") %>% 
    list.files(pattern = "clean_metadata", full.names = T) %>%
  file.info() %>% 
  dplyr::arrange(desc(ctime)) %>% 
  rownames() %>% 
  dplyr::first() %>% 
  read.csv()
  
mymre@metadata@categorical_vals <- here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@numeric_vals <- here::here("Metadata", "NumericalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@longitudinal_vals <- here::here("Metadata", "LongitudinalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

cat_vector <- get_cat(mymre) %>% 
  pull(CategoricalVariable)

num_vector <- 
  get_num(mymre) %>% 
  pull(NumericalVariable)


```

```{r support_funs, warning=F, message=F, echo = F}
peak_finder <-
  function(density, goal = "max",keep = F) {
    ### Let's define this function heere but we'll use it later
    stopifnot("goal can only be either local maxima (use: 'max') or local minima (use 'min)" = goal %in% c("min","max"))
    
    val <- case_when(goal == "max" ~ -2,
                   goal == "min" ~ 2)
    
    res <-
      density %>%
      purrr::keep(names(.) %in% c("x", "y")) %>%
      as.data.frame() %>%
      dplyr::mutate(is_peak = c(F, # 1st value can't be a local maxima, as there is no prior value to compare
                                diff(y) %>% # For each nth element of the vector, get (n+1) - n
                                  sign() %>%  # convert to -1 (n+1 is lower than n) or +1 (n+1 is higher than n)
                                  diff() == val, # if n is at a local maxima, both n+1 and n-1 are negative so [(n+1)-n] - [n - (n-1)] is -2
                                F)) # Last element can't be a local maxima either)
                    
                    if (!keep) {
                      res %>%
                        dplyr::filter(is_peak) %>%
                        dplyr::select(-is_peak)
                    } else{
                      res %>%
                        dplyr::select(-is_peak)
                    }
  }


```


```{r read_GR, echo = F, warning=F, message=F, results='hide'}
igc<-metar::get_diversity(mymre)@igc
Threshold <-  igc@dataTable %>%
  dplyr::select(SampleID, NumberMappedReads) %>%
  unique() %>%
  summarise(Q = quantile(NumberMappedReads,0.02)) %>% 
  as.numeric()


igc_df <- igc@dataTable %>%
                filter(NumberMappedReads >= Threshold) %>%
                filter(ReadCountReal>=Threshold) %>% ### Get the Mapped Read count (per-sample) just above the threshold
                group_by(SampleID) %>%
                summarise(across(everything(), min)) %>%
                ungroup() %>%
          right_join(., metadata, by="SampleID") %>%
          select(SampleID,
                 richness = GeneNumber,
                 link_var= record_id,
                 cat_var = group,
                 long_var = time_point,
                 NumberMappedReads, 
                 num_vector, 
                 height = Height,
                 weight = Weight) %>% 
  mutate(sCD14 = as.numeric(sCD14),
         weight = case_when(link_var == "C007" & long_var == 48 ~ 69.7,
                            link_var == "D009" & long_var == 96 ~ 61.9,
                            link_var == "E008" & long_var == 0 ~ 39.1,
                            link_var == "D013" & long_var == 96 ~ 67,
                            TRUE ~ as.numeric(weight)),
         BMI = weight/((height/100)^2))


```



## Stratification by BMI

One of the first things will be to test for how the BMI distributes at baseline, and to check for differences between groups


```{r basal_dist_df}
peaks_bl <- igc_df %>% 
  dplyr::filter(long_var == 0, !is.na(BMI)) %>% 
  dplyr::pull(BMI) %>%
  density() %>% 
  peak_finder(goal = "max")

median_bl <- igc_df %>% 
  dplyr::filter(long_var == 0, !is.na(BMI)) %>% 
  pull(BMI) %>% 
  median()

igc_df %>% 
  dplyr::filter(long_var == 0, !is.na(BMI)) %>% 
  ggplot(.) +
  # geom_histogram(aes(x = BMI), fill = "grey", bins = 100) +
  geom_density(aes(x = BMI),bins = 100, fill = "grey") +
  theme_bw() +
  geom_vline(xintercept = peaks_bl$x, lty = 3) +
  geom_vline(xintercept = median_bl, col = "red") +
  geom_text(data = peaks_bl, aes(x = x, y = y*.9, label = round(x, 3))) +
  geom_text(aes(x = median_bl, y = max(peaks_bl$y)*.5, label = round(median_bl,3))) +
  labs(x = "BMI", y = "kernel density", title = "BMI distribution at baseline") +
  theme(axis.title.y = element_blank())



igc_df %>% 
  dplyr::filter(long_var == 0, !is.na(BMI)) %>% 
  dplyr::mutate(BMI_c = case_when(BMI >= 22 ~ "high",
                                  BMI < 22 ~ "low")) %>% 
  group_by(BMI_c) %>% 
  tally() %>% 
  kableExtra::kable(format="markdown")









```

It seems there is a certain bimodality in the distribution of BMI at BL. However, still seems to be close to normal, so the max peak of 21.57 is fairly close to the median of 22.084. From this it 22 seems a good cut-off point between high and low BMI.



```{r kmeans_split, warning=FALSE, message=FALSE}


bmi_c <- 
igc_df %>% 
  dplyr::filter(long_var == 0, !is.na(BMI)) %>% 
  dplyr::select(link_var, long_var, cat_var, BMI, richness) %>% 
  mutate(cluster = kmeans(BMI, centers = 2)$cluster)

bmi_c %>% 
  arrange(BMI) %>% 
  filter(BMI == max(BMI[cluster == 1]))

```

A kmeans clustering approach splits both groups at around BMI = 22.5. This should further reiforce the 22 threshold.

# STratification by BMI at baseline

Now we found a threshold appropiate to split by BMI, we can stratify the population by treatment and bmi, vs gene richness.


```{r BL_stratified_analysis, warning = F, message = F}

bmi_c %>% 
  dplyr::mutate(BMI_c = case_when(
    BMI >= 22 ~ "high",
    BMI < 22 ~ "low"
  )) %>% 
  
  # dplyr::mutate(group_bmi = paste(cat_var, BMI_c)) %>% 
  ggplot(aes(x = cat_var, y = richness, fill = BMI_c)) +
  geom_boxplot() + 
  ggpubr::stat_compare_means(method = "wilcox.test", hide.ns = F) +
  theme_light()


bmi_c %>% 
  # dplyr::select(BMI, richness) %>% 
  dplyr::filter(!is.na(richness)) %>% 
  mutate(new_c = kmeans(.[,c("BMI","richness")], centers = 2)$cluster) %>% 
  ggplot(aes( x = BMI, y = richness, color = cat_var)) +
  geom_point() +
  theme_light()

```
Doesn't seem to be any noticeable difference between groups or BMI types. No diferences between groups could be found, 


# Stratification by BMI changes

Maybe the increase in gene richness may be associated with higher increase in BMI

```{r, eval = F}

delta_bmi_df
igc_df %>% 
  # dplyr::filter(long_var == 0) %>% 
  dplyr::select(link_var, long_var, cat_var, richness, BMI) %>% 
  group_by(link_var) %>% 
  dplyr::filter(long_var %in% c(0,96), 
                !is.na(BMI)) %>% 
  filter(n() ==2) %>% 
  mutate(delta_BMI_abs = BMI - BMI[long_var == 0],
         delta_BMI_rel = (BMI - BMI[long_var == 0])*100 / BMI[long_var == 0],
         delta_gr_abs = richness - richness[long_var == 0],
         delta_gr_rel = (richness - richness[long_var == 0]) / (richness + richness[long_var == 0])) %>%
  dplyr::filter(long_var == 96) 



%>%
  ggplot(aes(x = delta_BMI_abs)) +
  geom_density()
  
  
  create_LMM(.,link_var = "link_var", cat_var = "cat_var",long_var = "delta_gr_rel",num_var = "delta_BMI_rel",breakpoints = NULL)


```

