---
title: "LMM for taxa"
author: "Marc Noguera"
date: "3/29/2022"
output: html_document
---

```{r setup, include=T,eval=F}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

```

```{r utils, message=F, warning=F}
get_LMM_GraphParams <- function(Table) {
  
  myList <- Table %>%
    filter(!is.na(cat_var)) %>%
    pull(cat_var) %>%
    unique() %>%
    set_names() %>%
    map_dfr(~ {
      
      innerTable <- Table %>%
        dplyr::select(num_var, long_var, cat_var, id_var) %>%
        dplyr::filter(cat_var == .x) 
      
      if(sum(innerTable$num_var) == 0){  ## Check num var to not be all 0s, otherwise the lme function will fall apart.
        
        graphparams <- Table %>%
        summarise(
          xmin = min(long_var),
          xmax = max(long_var),
          ymin = 0,
          ymax = 0
        ) %>%
        mutate(cat_var = .x) %>%
        relocate(cat_var)


        
      } else if(length(unique(innerTable$id_var)) < 2){
        
        graphparams <- Table %>%
        summarise(
          xmin = min(long_var),
          xmax = max(long_var),
          ymin = min(num_var),
          ymax = max(num_var)
        ) %>%
        mutate(cat_var = .x) %>%
        relocate(cat_var)

      } else {      
      
        model <- 
        lmerTest::lmer(formula = "num_var ~ long_var + (1|id_var)", innerTable) %>%
        summary() %>%
        magrittr::extract2("coefficients")
      
      graphparams <- Table %>%
        summarise(
          xmin = min(long_var),
          xmax = max(long_var),
          ymin = model[1, 1] + xmin * model[2, 1],
          ymax = model[1, 1] + xmax * model[2, 1]
        ) %>%
        mutate(cat_var = .x) %>%
        relocate(cat_var)

      }
      
    }) %>%
    as.data.frame()
  
  
  return(myList)
}





```



##  Taxa level LMM

The main purpose of this script is to extract signals from Linear-Mixed-Models(LMM) from longitudinal designs of data. For each Taxa, at a tax_level, we'll run a LMM through the longitudinal dimension and either, rank taxa according to their longitudinal increase/decrease or to the highest differences between a group variable in cat_var.

Try different data transformation approaches:

* Relative abudnaces(RA)
* raw counts, with previous rarefaction at a certain number of reads, using metaphlan_counts object
* clr transformed data for counts, using microbiome::transform()

```{r extract data, echo=T, include=T, warning=F, message=F}
require(phyloseq)
require(ggstatsplot)
require(ggplot2)
require(here)
library(metar)
library(tidyverse)

bucket<-"s3://mistral-wp6-advanz4/"
bucketDir<-"metagenome/WMGS"

# Load package from local copy

devtools::load_all(here::here("..","WMGSPipeline"))### Change to require/library/load local package

mymre <- readRDS(here::here("data", "mre.rds"))
# mymre<-aws.s3::s3readRDS(bucket=bucket,object = paste0(bucketDir,"/MREObject.rds")) ## Change as needed
# mymre@metadata@categorical_vals <-
#   here::here("Metadata","CategoricalVariables.txt") %>%
#   read.table(sep="\t") %>%
#   setNames(c("CategoricalVariable","PaletteName")) %>%
#   as_tibble()
phy <- metar::get_phyloseq(mymre, type="metaphlan")
id_var <- "record_id"
long_var <- "time_point"
cat_df <- metar::get_cat(mymre) %>%
  filter(!CategoricalVariable == "CD4diff_96")
cat_vector <- cat_df$CategoricalVariable

metadata <-
  metar::get_meta(mymre) %>%
  mutate(rnames = SampleID) %>%
  tibble::column_to_rownames("rnames")

## Remove patients with <2 samples, this is temporal 

selectedIds <- metadata %>%
  select(SampleID, record_id) %>%
  group_by(record_id) %>% 
  summarise(n = n()) %>% 
  filter(n >= 2) %>% 
  pull(record_id)

metadata <- metadata %>%
  filter(record_id %in% selectedIds)

abundance <- 
  metar:::get_merged_df(phy, tax_level="Species",sample_data_df = F) %>%
  full_join(
  metadata[,c("SampleID",id_var,long_var, cat_vector)], by="SampleID") %>%
  dplyr::rename(id_var = !!sym(id_var),
                long_var = !!sym(long_var))


  



```

## Including Plots



```{r LMM_effect_table, warning=F, message=F}


num_vector <- abundance %>% 
  dplyr::select(-id_var, -cat_vector, -long_var, -SampleID) %>%
  colnames()




get_multiLMM_table <- function(df,num_vector, sam_var, cat_vector, long_var, id_var){
  
  myUnifiedModelList <- cat_vector %>%
  set_names() %>%
  
  map(function(cat_var) {
    # print(cat_var)
    num_vector %>%
      set_names() %>%
      
      map_dfr(function(num_var) {
        
        dat <-
          df %>%
          select(sam_var, id_var, long_var, cat_var, num_var) %>%
          dplyr::rename(num_var = !!sym(num_var),
                        cat_var = !!sym(cat_var))
        
        
        cat_levels <- #Count nº of levels in cat_var
          dat %>% 
          dplyr::select(id_var, cat_var) %>%
          unique() %>%
          dplyr::group_by(cat_var) %>%
          dplyr::summarize(n=n()) %>%
          dplyr::filter(n>1,
                        !is.na(cat_var)) %>%
          dplyr::pull(cat_var) 
        

        dat <- # filter NAs and bad entries
          dat %>%
          dplyr::filter(cat_var %in% cat_levels,
                 !is.na(cat_var))
        
        
        if (length(!is.na(cat_levels)) >= 2){ # Check if cat_var has at least 2 levels
        
        myUnifiedModel <- 
          lmerTest::lmer(formula = "num_var ~ long_var * cat_var + (1|id_var)", data = dat)
        
        finalTable <-
          anova(myUnifiedModel) %>%

          # magrittr::extract2("coefficients") %>%
          as.data.frame() %>%
          tibble::rownames_to_column(var = "term") %>%
          dplyr::filter(term == "long_var:cat_var") %>%
          magrittr::set_names(c("term","SumSq", "MeanSq", "df", "DenDF", "Fval","pval")) %>%
          dplyr::mutate(pval = round(pval, 3)) %>%
          mutate(term = num_var) %>%
          rename(Species = term)
        

        return(finalTable)
        
        } else {return(NA)}
      })
    
  })

}


taxa_ranks <- get_multiLMM_table(df = abundance,
                                num_vector = num_vector,
                                sam_var = "SampleID",
                                cat_vector = cat_vector, 
                                long_var = long_var,
                                id_var = id_var)


signifDF <- taxa_ranks$group %>% 
  arrange(desc(Fval)) %>%
  mutate(padj = p.adjust(pval, method= "fdr")) %>%
  filter(pval < 0.01)


```
As this was a proper LMM using cat_var as fixed effect, we can't get an overall "slope", as the model computes an estimate for each cat_var level. Hence, as a proxy for effect size we'll ANOVA F-statistic which is calculated as the division betwen inter-group mean variance and intra-group variance. From the significant species we'll be able to greatly reduce the table so we can comfortably estimate slopes for each one.

By looking at group test statistic, we see a few species with a significant time:group interaction, which becomes unsignificant even with a permisive fdr correction. 

```{r slopes_and_plots, warning=F, message=F}

taxaRank_plotlists <- function(rankdf, otus, cat_var){
  rankdf %>%
  pull(Species) %>% 
  set_names() %>%
  map(function(sp) {
    
   sp
    
    dat <-
      otus %>% 
      dplyr::filter(!is.na(!!sym(cat_var))) %>%
      select(num_var = sp,
             cat_var = !!sym(cat_var),
             id_var,
             long_var)
    
    model <- dat %>%
      lmerTest::lmer(formula = "num_var ~ long_var * cat_var + (1|id_var)", data = .)
    
    myGraphParams <- dat %>%
      get_LMM_GraphParams() %>%
      mutate(across(contains("x"), ~ as.numeric(.x)))
    
    myPval <-
      anova(model) %>%
      
      # magrittr::extract2("coefficients") %>%
      as.data.frame() %>%
      dplyr::slice(3L) %>%
      magrittr::set_names(c("SumSq", "MeanSq", "df", "DenDF", "Fval", "pval")) %>%
      dplyr::mutate(pval = round(pval, 3))
    
    myplot <- dat %>%
      ggplot(.,
             aes(
               x = long_var,
               y = num_var,
               group = id_var,
               color = cat_var
             )) +
      geom_point(alpha = 0.2) +
      geom_line(alpha = 0.2) +
      scale_x_continuous(breaks = unique(dat$long_var),
                         limits = c(min(dat$long_var),
                                    max(dat$long_var))) +
      
      geom_segment(
        data = myGraphParams,
        aes(
          x = xmin,
          y = ymin,
          xend = xmax,
          yend = ymax,
          color = cat_var
        ),
        inherit.aes = T,
        size = 2
      ) +
      scale_color_brewer(palette = "Set1") +
      labs(color = cat_var,
           x = "weeks",
           y = sp) +
      geom_label(
        data = myPval,
        aes(
          alpha = NULL,
          label = paste("ANOVA p=",
                        pval,
                        sep = ""),
          x = -Inf,
          y = Inf
        ),
        color = "black",
        hjust = 0.01,
        vjust = 1
      ) +
      theme_bw()
    
  })
}


plot_list <- cat_vector %>%
  set_names() %>%
  map(function(cat) {  
    cat_var <- cat
    taxaRank_plotlists(rankdf = signifDF,otus = abundance, cat_var = cat)
})


plot_list$group

```

