---
title: "Taxonomy: Differential Abundance"
author: "Carlos Blázquez Bondia"
date: "3/23/2022"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RColorBrewer)
library(metar)

```

## Data import and exploration

This section should explain anyithing that may have happened during the first fase of importing and formatting data, such as any modification or additional filtering step.

```{r Setup, warning=F, message=F}

# mymre <- here::here("data","mre.rds") %>%
#   readRDS(.)

mymre <- aws.s3::s3readRDS(bucket = "s3://mistral-wp6-advanz4", object = "metagenome/WMGS/MREObject.rds")
# metadata <-get_meta(mymre)
metadata <- here::here("Metadata", "2022_04_28_clean_metadata_LIMS.csv") %>% 
  read.csv() %>% 
  column_to_rownames("SampleID") %>% 
  mutate(SampleID = rownames(.))



cat_df <- 
  here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T)

cat_vector <- cat_df %>% 
  pull(CategoricalVariable)

long_var <- here::here("Metadata", "LongitudinalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(LongitudinalVariable)

link_var <- here::here("Metadata", "LongitudinalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(LinkVariable)

num_var <- here::here("Metadata", "NumericalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(NumericalVariable)



```



```{r update_mre, warning= F, message= F}

#### This chunk is for updating the mre internally as new variables come out. Once the final mre is done this chunk will be removed.

devtools::load_all(here::here("../WMGSPipeline"))

mymre <- filter_samples(mre = mymre, sample_ids = metadata$SampleID)

mymre@metadata@metadata_df <- as.tibble(metadata) 
mymre@taxa@metaphlan@phyloseq@sam_data <- metadata %>%
  phyloseq::sample_data(.)
mymre@taxa@metaphlan@phyloseq_sec@sam_data <- metadata %>%
  phyloseq::sample_data(.)




mymre@metadata@categorical_vals <- here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@numeric_vals <- here::here("Metadata", "NumericalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@longitudinal_vals <- here::here("Metadata", "LongitudinalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()



```


```{r declare_support_functions}
# Any homebrew support function that may be written will be placed here

```


First we'll start with a non-compositional approach which may provide potential candidate taxa for DA to be confirmed with more astringent and specific compositional approaches.


# Non-compositional analysis

In this case, we'll test for group as categorical variables. We may extend the analysis towards other cat_vars, but those were more important for basal comparisons, which already shown no great differences were found among other potential confusor variables. 

## Cross-group Differential Abundance

These comparisons will be between treatment groups at the same time point: one at week 0, week 48 and 96.
```{r wilcox_across_group, warning=F, message=F}

phy <- get_phyloseq(mymre, type = "metaphlan") %>% 
      phyloseq::filter_taxa(., function(x) sum(x > 0) > (0.3*length(x)), TRUE)


otu_df <- 
  phyloseq::otu_table(phy) %>% 
  t() %>% 
  as.data.frame() %>% 
  stats::setNames(phyloseq::tax_table(phy)[,"Species"]) %>% 
  tibble::rownames_to_column("SampleID") %>% 
  stats::setNames(.,stringr::str_remove_all(colnames(.), "\\[")) %>% 
  stats::setNames(.,stringr::str_remove_all(colnames(.), "\\]"))


stat_list <-
  cat_vector %>%
  purrr::set_names() %>%
  purrr::map(function(cv) {
    
    c(0,48,96) %>% 
      set_names() %>% 
      map_dfr(function(tp){
        otu_df %>%
        dplyr::select(-SampleID) %>%
        colnames() %>%
        purrr::set_names() %>%
        purrr::map_dfr(function(sp) {
          df <-
            otu_df %>%
            dplyr::select(SampleID, !!sym(sp)) %>%
            dplyr::left_join(metadata[, c("SampleID", link_var, cv, long_var), drop = F], 
                             by = "SampleID") %>% 
            dplyr::filter(!!sym(long_var) %in% tp) %>% 
            dplyr::group_by(!!sym(long_var)) 

          
          if (length(unique(pull(.data = df, !!sym(cv)))) == 2) {
            test <- df %>%
              rstatix::wilcox_test(., formula(paste(sp, "~", cv)), paired = F) %>%
              dplyr::mutate(method = "Wilcoxon")
            
          } else{
            test <- df %>%
              rstatix::kruskal_test(., formula(paste(sp, "~", cv))) %>%
              dplyr::mutate(method = "Kruskal-Wallis")
          }
          
        }) %>% 
      rename(Species = .y.) %>% 
      group_by(!!sym(long_var)) %>% 
      mutate(p.adj = p.adjust(p, "BH")) %>% 
      # filter(p.adj < .05 ) %>% 
      group_by(Species) 
      }) 
}) 


DAsp <- stat_list$group %>% 
  pull(Species)

## Most differences are found on w48 and w96.
## Produce a table showing 
stat_list$group %>% 
  filter(p<0.05) %>% 
  group_by(Species) %>% 
  tally() %>% 
  filter(n>=1) ## Species being significant (Wilcoxon unadjusted <0.05)

### Show how many species are differential in at least two of the timepoints in each cat var. Risk_group should result in many species showing that kins of signal. See Venn section below.



otu_df %>% 
  select(SampleID, !!sym(DAsp)) %>% 
  left_join(., metadata[,c("SampleID","group","time_point")], by="SampleID") %>% 
  group_by(group) %>% 
  ggstatsplot::grouped_ggbetweenstats(.,
                              x="group",
                              y = !!sym(DAsp[1]),
                              pairwise.comparisons = T,
                              ylab = paste(DAsp[1], "(%)"), 
                              ggtheme = ggplot2::theme_bw(),
                              grouping.var = "time_point",plotgrid.args = NULL)


```

Only one spec


### Comparison of significant species

Following Venn's diagram to detect group_specific DA taxa (Marc's addition, moved here as I thought it made more sense structurally)
```{r wilcox_Venn_Group, warning=F, message=F}

```

## Longitudinal, within-group Differential Abundance
These comparisons test differences between time points within the same group. This is a longitudinal approach so it requires a paired test, which account for dependece between data points.
```{r wilcox_within_group, message=F, warning=F}
            # dplyr::filter(!n()<2) %>% 
            # dplyr::arrange(!!sym(link_var), !!sym(long_var))
```


# Compositional approach (ANCOM)

## Cross-group Differential Abundance

These comparisons will be between treatment groups at the same time point: one at week 0, week 24, week 48 and 96.
```{r ANCOM_across_group, warning=F, message=F}

```

### Comparison of significant species

Following Venn's diagram to detect group_specific DA taxa (Marc's addition, moved here as I thought it made more sense structurally)
```{r ANCOM_Venn_Group, warning=F, message=F}

```

## Longitudinal, within-group Differential Abundance
These comparisons test differences between time points within the same group. This is a longitudinal approach so it requires a paired test, which account for dependece between data points.

```{r ANCOM_within_group, warning=F, message= F}

```

