---
title: "Taxonomy: Differential Abundance"
author: "Carlos Blázquez Bondia"
date: "3/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Data import and exploration

This section should explain anyithing that may have happened during the first fase of importing and formatting data, such as any modification or additional filtering step.

```{r read_data, eval=F}
bucket <- "s3://mistral-wp6-advanz4/"
bucket_dir <- "metagenome/WMGS/"

# mymre <- aws.s3::s3readRDS(bucket = bucket, object = paste0(bucket_dir,"MREObject.rds"))
mymre <- here::here("data","mre.rds") %>%
  readRDS() 

```


```{r declare_support_functions}
# Any homebrew support function that may be written will be placed here

```


# Differential abundance results

Here we will leave all results regarding DA. In case we analyze it by more than one approach (e.g, wilcoxon, ANCOM, etc), each will have it's own subsection)

## Method 1 (e.g: Wilcoxon on count data)

## Method 2 (e.g compositional approach)


## Method 3 (Marc Addition), test differential abundance between groups at each timepoint. Create Venn diagram of tiempoint-specific DA taxa.


```{r pressure, echo=FALSE}

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
