---
title: "Species vs inflammation"
author: "Carlos Blázquez Bondia"
date: '2022-05-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(RColorBrewer)
library(metar)
source("code/LMM.R")
source("code/myFunctions.R")

```

## Data import and exploration

This section should explain anyithing that may have happened during the first fase of importing and formatting data, such as any modification or additional filtering step.

```{r Setup, warning=F, message=F}

# mymre <- here::here("data","mre.rds") %>%
#   readRDS(.)

mymre <- aws.s3::s3readRDS(bucket = "s3://mistral-wp6-advanz4", object = "metagenome/WMGS/MREObject.rds")
# metadata <-get_meta(mymre)
metadata <- here::here("Metadata", "2022_04_28_clean_metadata_LIMS.csv") %>% 
  read.csv() %>% 
  column_to_rownames("SampleID") %>% 
  mutate(SampleID = rownames(.))



cat_df <- 
  here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T)

cat_vector <- cat_df %>% 
  pull(CategoricalVariable)

long_var <- here::here("Metadata", "LongitudinalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(LongitudinalVariable)

link_var <- here::here("Metadata", "LongitudinalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(LinkVariable)

num_vector <- here::here("Metadata", "NumericalVariables.txt") %>%
  read.delim(., header = T)  %>% 
  pull(NumericalVariable)



```



```{r update_mre, warning= F, message= F}

#### This chunk is for updating the mre internally as new variables come out. Once the final mre is done this chunk will be removed.

devtools::load_all(here::here("../WMGSPipeline"))

mymre <- filter_samples(mre = mymre, sample_ids = metadata$SampleID)

mymre@metadata@metadata_df <- as.tibble(metadata) 
mymre@taxa@metaphlan@phyloseq@sam_data <- metadata %>%
  phyloseq::sample_data(.)
mymre@taxa@metaphlan@phyloseq_sec@sam_data <- metadata %>%
  phyloseq::sample_data(.)




mymre@metadata@categorical_vals <- here::here("Metadata", "CategoricalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@numeric_vals <- here::here("Metadata", "NumericalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()

mymre@metadata@longitudinal_vals <- here::here("Metadata", "LongitudinalVariables.txt") %>% 
  read.delim(., header = T) %>% 
  tibble()



```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
phy <- get_phyloseq(mymre, type = "metaphlan") %>% 
      phyloseq::filter_taxa(., function(x) sum(x > 0) > (0.3*length(x)), TRUE)

num_vector <- metar::get_num(mymre) %>% 
  dplyr::pull(NumericalVariable)
strep_df <- 
  phyloseq::otu_table(phy) %>% 
  t() %>% 
  as.data.frame() %>% 
  stats::setNames(phyloseq::tax_table(phy)[,"Species"]) %>% 
  tibble::rownames_to_column("SampleID") %>% 
  stats::setNames(.,stringr::str_remove_all(colnames(.), "\\[")) %>% 
  stats::setNames(.,stringr::str_remove_all(colnames(.), "\\]")) %>% 
  dplyr::select(SampleID,species = Streptococcus_parasanguinis) %>% 
  dplyr::inner_join(metadata[,c("SampleID", link_var,long_var,cat_vector, num_vector)])


num_vector %>% 
  purrr::set_names() %>% 
  purrr::map(function(nv){
    strep_df %>% 
      dplyr::select(cat_var = group,
                    long_var = time_point,
                    link_var = link_var,
                    species,
                    num_var = !!sym(nv)) %>% 
      dplyr::filter(!is.na(species)) %>% 
      dplyr::mutate(species = as.numeric(species))
      create_LMM(., num_var = "species",
                 long_var = "num_var", 
                 cat_var = "cat_var", 
                 breakpoints = NULL,
                 link_var = "link_var" )
  })




basal <-
  metadata %>%
  filter(time_point == 0) %>%
  pull(record_id) %>%
  unique()

thresholds <- data.frame(
  num_var = num_vector,
  threshold = c(150,50,30, )
)


num_vector %>%
  purrr::set_names() %>%
  purrr::map(
    function(nv) {

      
      metadata %>%
        dplyr::filter(record_id %in% basal) %>%
        dplyr::select(record_id, group, num_var = nv, time_point) %>%
        dplyr::group_by(record_id) %>%
        dplyr::mutate(basal_num = num_var[time_point == 0]) %>%
        dplyr::mutate(diff = num_var - basal_num) %>%
        dplyr::mutate(thres = case_when(diff >= 150 ~ "yes", diff < 50 ~ "no")) %>%
        dplyr::select(-basal_num) %>%
        dplyr::group_by(group, time_point) %>%
        dplyr::filter(!is.na(thres)) %>%
        dplyr::count(thres) %>%
        dplyr::mutate(sumn = n / sum(n)) %>%
        dplyr::ungroup() %>%
        dplyr::mutate(time_point = as.factor(time_point)) %>%
        ggplot(aes(x = time_point, y = sumn, fill = thres)) +
        facet_wrap( ~ group) +
        geom_bar(stat = "identity", position = "stack") +
        scale_fill_brewer(palette = "Set1") +
        theme_bw() +
        labs(x = "week", y = "percentage", title = nv)
    })


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
