---
title: "Diversity_GeneRichness"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(metar)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
library(lme4)
library(lmerTest)
source(file = "code/myFunctions.R")

```




```{r load_metadata_vars, eval=T, include=F}
myPath<-here::here()
setwd(myPath)
mymre<-readRDS("data/mre.rds")

metadata <- metar::get_metadata(mymre)@metadata_df


CategoricalVariablesDF <- get_metadata(mymre)@categorical_vals
NumericalVariablesDF <- get_metadata(mymre)@numeric_vals
LongitudinalVariablesDF <- get_metadata(mymre)@longitudinal_vals





```



```{r get_GeneRichness_Data, echo=FALSE, message=F, warning=F}


igc<-metar::get_diversity(mymre)@igc

# We'll save this number to report the minimum read threshold.
Threshold = igc@dataTable %>%
  summarise(Q = quantile(NumberMappedReads,0.02)) %>%
  as.numeric()

ThresholdTable<-igc@dataTable %>%
  filter(NumberMappedReads >= Threshold) %>%
  group_by(SampleID) %>%
  summarise(across(everything(), min)) %>%
  ungroup() %>%
  left_join(., metadata, by="SampleID") %>%
  rename(richness = GeneNumber,
         link_var=record_id,
         cat_var = Group,
         long_var = TimePoint)
  


```

## Longitudinal evolution: LMMs



```{r create_LMM_Function, echo=FALSE}



create_LMM <- function(metadata, num_var, cat_var, long_var, link_var,breakpoints){
  
  myLMM <-metadata %>%
    as.data.frame() %>%
    dplyr::select(link_var = !!sym(link_var),
                  num_var = !!sym(num_var),
                  cat_var = !!sym(cat_var),
                  long_var = !!sym(long_var)) %>%
    dplyr::mutate(cat_var = as.factor(cat_var)) %>%
    myLMM_two.piece.GLMM(Datos = .,
                         Y = "num_var",
                         Time = "long_var",
                         Factor = "cat_var",
                         t = breakpoints,
                         ID = "link_var")

  
  myLMM
}


get_LMM_GraphParams <- function(Table){
  
    myList <- Table %>%
    pull(cat_var) %>%
    levels() %>%
    set_names() %>%
    map_dfr( ~ {
      
    model<-Table %>%
        dplyr::select(num_var, long_var, cat_var, link_var) %>%
        dplyr::filter(cat_var == .x) %>%
        lmerTest::lmer(formula = "num_var ~ long_var + (1|link_var)") %>%
        summary() %>%
        magrittr::extract2("coefficients") 
    
    graphparams<- ThresholdTable %>%
      summarise(xmin=min(long_var),
                xmax=max(long_var),
                ymin = model[1,1] + xmin*model[2,1],
                ymax = model[1,1] + xmax*model[2,1]) %>%
      mutate(cat_var = .x) %>%
      relocate(cat_var)
  }) %>% 
      as.data.frame() 
  
    
    return(myList)
}


```




```{r plot_LMMs, message = F, warning = F}

cat_var<-CategoricalVariablesDF$CategoricalVariable
long_var <- LongitudinalVariablesDF$LongitudinalVariable[1]
link_var<-LongitudinalVariablesDF$LinkVariable[1]
num_var <- "richness"
t<-NULL
# 
plot_ist<-create_LMM(metadata = ThresholdTable,
           num_var = "richness", 
           cat_var = "cat_var",
           long_var = "long_var",
           link_var = "link_var",
           breakpoints = NULL)

plot_list <-  c(num_var) %>%
  purrr::set_names() %>%
  purrr::map( ~ {
    LMM<- create_LMM(ThresholdTable, 
               num_var = .x, 
               cat_var = "cat_var",
               long_var = "long_var",
               link_var = "link_var",
               breakpoints = t)
    
    pvals<-names(LMM[[1]]) %>%
      set_names() %>%
      
      map_dfr(~ {
        LMM[[1]][[.x]] %>%
          as.data.frame() %>%
          slice(2) %>%
          select("p-value", "Value") %>%
          as.numeric() %>%
          round(digits=3)
        }) %>% 
      
      t() %>%
      as.data.frame() %>%
      setNames(., nm= c("pval", "slope")) %>%
      rownames_to_column(var="cat_var")
    
    
      LMM[[2]] +
      geom_label(data=pvals, 
                 aes(alpha=NULL,label=paste("p=",pval, ", slope = ", slope, sep=""), 
                     x=-Inf, y=Inf), color="black", hjust=0.01, vjust=1 ) +
      labs(title=.x, color = cat_var)

  }
)


ggarrange(plotlist = plot_list, common.legend = T)
```

There is a pretty clear evolution in every one of the treatment groups. However the evolution rate seems pretty similar between both treatment groups. The models fail to find a great difference in gene richness evolution, althougha slight yet significant increasing trend (slope value = 0.027), can be found in the DGT group (Group 1).

However, creating a single LMM with treatment group as a random effect whe can meassure the actual difference in the effect of timepoint over gene richness between two groups. 


```{r compare_group_effect}


myUnifiedModel<- ThresholdTable %>%
  dplyr::rename(num_var = richness) %>%
  lmerTest::lmer(formula = "num_var ~ long_var * cat_var + (1|link_var)",data = .)

myGraphParams<-ThresholdTable %>%
  dplyr::rename(num_var = richness) %>%
  get_LMM_GraphParams() %>%
  mutate(across(contains("x"), ~ as.numeric(.x)))

myPval<-summary(myUnifiedModel) %>%
  magrittr::extract2("coefficients") %>%
  as.data.frame()%>%
  slice(4L) %>%
  set_names(c("Estimate", "Std.Err", "df", "t.val","pval")) %>%
  mutate(pval = round(pval,3))

ThresholdTable %>%
  dplyr::rename(num_var = richness) %>%
  ggplot(., aes(x =long_var, y=num_var, group=link_var, color=cat_var)) +
  geom_point(alpha=0.4) + 
  geom_line(alpha=0.4) + 
  scale_x_continuous(breaks = unique(ThresholdTable$long_var),
                     limits = c(min(ThresholdTable$long_var),
                                max(ThresholdTable$long_var))) +

  geom_segment(data=myGraphParams,aes(x=xmin,y=ymin,xend=xmax,yend=ymax, color=cat_var),inherit.aes=T,size=3) +
  scale_color_brewer(palette="Set1") +
  labs(color="Treatment", x="weeks") +
  geom_label(data=myPval, 
             aes(alpha=NULL,label=paste("ANOVA p=",pval, sep=""),
                 x=-Inf, y=Inf), color="black", hjust=0.01, vjust=1 ) +
  theme_bw() 
  
summary(myUnifiedModel) %>%
  magrittr::extract2("coefficients") %>%
  as.data.frame() %>% kableExtra::kable(., format="markdown")


```


This shows a slight effect of treatment, where the effect of time over richness in group group2 is 0.025 (St.err = 0.025) times lower than the same for group 1 (ANOVA p=0.31)


```{r boxplots, eval=F, echo=F, message=F}


### This was a first test to learn how to perform multiple comparisons in a true tidyverse fashion

plot_list <-  NumericalVariablesDF %>%
  
  dplyr::pull(NumericalVariable) %>%
  purrr::set_names() %>%
  purrr::map( ~ {
    metadata %>%
      dplyr::select(cat_var = !!sym(cat_var),
                    long_var = !!sym(long_var),
                    num_var = .x,
                    link_var = !!sym(link_var)) %>%
      ggplot(., aes(x = as.factor(long_var), y = num_var, fill = cat_var)) +
      geom_boxplot() +
      stat_compare_means(method="wilcox.test", 
                         label = "p.signif", 
                         hide.ns = T, bracket.size = .6, 
                         label.y.npc ="center") +
      labs(title = .x,
          x = "Week",
          y="counts",
          fill=cat_var) +
      theme_bw() + 
      scale_fill_brewer(palette = CategoricalVariablesDF$PaletteName[1])
  }
)

ggarrange(plotlist = plot_list, common.legend = T)



```
