---
title: "TODO"
author: "Marc Noguera"
date: "2/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##### TO - DO (try to use metar package )

Carlos, en aquest document vaig especificant els análisis targeted a aquest estudi que hauries d'anar treballant. L'estructura de workflowr i com l'has fet servir sta molt bé. Et recomano que facis servir el paquet metar quan puguis.

Compare, both through timepoint-specific boxplots and LMM how bray-curtis distance vs baselins has changes in each treatment-group and in each Delta-CD4 group as defined in analysis plan
```{r Bray-Curtis Distance vs baseline}

```


perform PAM analysis using baseline samples only and describe differences
```{r baseling PAM-clustering analysis}
```

Using Delta-CD4 grouping, check if baseline and week 96 microbiome is different
```{r Delta-CD4 group vs baseline}
```

Using Delta-CD4 grouping and baseling clinical data only, check from clinical differences between group. Use, compareGroups.

